MATH 245 Homework 4

### 2024-04-03 PDEs/math245_hw4.Rmd
finished pde 4

---
title: "MATH 245 Homework 4"
author: "Ruby Krasnow and Tommy Thach"
date: "2024-04-01"
output:
  pdf_document:
    extra_dependencies: ["tcolorbox", "xcolor"]
---

@@

Re-writing as $X''+\lambda X=0$, we will get the characteristic equation: $r^2+\beta^2=0$. Since our characteristic equation has complex roots $r= \pm i\beta$, our solutions take the form 
$$X(x)=A\sin{(\beta x)}+B\cos{(\beta x)}$$
Differentiating, we find that 
$$X'(x)=\beta A\cos{(\beta x)}-\beta B\sin{(\beta x)}$$
Now plugging in our initial condition $X'(0) = 0$, we get
$X'(0)=\beta A=0$. And since we are in a case where $\beta \neq 0$, this means $B=0$ so $X(x)=A\sin{(\beta x)}$. Now we use our boundary condition $X(l)=0$ to get $X(l)=A\sin{(\beta l)}=0$. If $A=0$, then $X(0)=0$ and this contradicts the definition of an eigenfunction. Therefore, we must have $\sin{(\beta l)}=0$, which can only occur if $\beta = \frac{n\pi}{l}$. Therefore, this case gives us eigenvalues

$$\lambda_n= \left(\frac{n\pi}{l}\right)^2 \quad n=1,2,3... $$
with eigenfunctions
$$X_n(x)=\sin{\left( \frac{n \pi x}{l}\right)}$$

@@

\underline{Case 2:} Zero eigenvalues, $\lambda=0$
$X''=0$ implies that $X(x)$ is of the form $Ax+B$. Now plugging in our initial condition $X(1) = 0$, we get $A+B=0$, which means $X(x)=Ax-A$. But from the boundary condition $X(e)=0$, we get $A(e-1)=0$ which is only possible if $A=0$ and accordingly $X(0)=0$. Therefore, there are no eigenfunctions $X(x)$ that satisfy the boundary conditions when $X''=0$ and hence no zero eigenvalues.

@@

$$X(1)=A+B=0 \quad \longrightarrow\quad A=-B$$
$$X(x)=Ax^{\beta }-Ax^{-\beta}$$
Then the other boundary condition gives $X(e)=Ae^{\beta}-Ae^{-\beta }=0$. Since $e^{\beta}$ and $e^{-\beta}$ are always nonzero, we must have $A=0$, implying that again $X(x)=0$.Thus, this problem has only positive eigenvalues and their associated eigenfunctions as found in Case 1.

@@

$$X(1)=A\sin{(\beta)}+B\cos{(\beta)}=0$$
$$A\sin{(\beta)}=-B\cos{(\beta)}$$

@@

$$ X'(0) + X(0) = \beta A-\beta B+A+B$$
$$0=A(\beta+1)+B(1-\beta)$$
$$-A(\beta+1)=B(1-\beta)$$
$$\frac{\beta+1}{1-\beta}=-\frac{B}{A}$$
Now turning to our other boundary condition,
$$X'(1)=\beta Ae^{\beta}-\beta Be^{-\beta}=0$$
$$\beta Ae^{\beta}=\beta Be^{-\beta}$$
$$-\frac{B}{A} = -e^{2\beta}$$
This would give us the following equation:
$$\frac{\beta+1}{1-\beta}=-e^{2\beta}$$
But the only time that this equality holds true is when $\beta=0$ (Fig. 2), but we are in the case when we defined $\beta$ to be nonzero, meaning that we have reached a contradiction and we cannot have any negative eigenvalues.

@@

Here, $l=\pi$. Let's start with $a_0$:
\begin{align*}
a_0 &=\frac{1}{2\pi} \int_{-\pi}^{\pi}f(x)dx \\
    &=\frac{1}{2\pi}\left[ \int_{-\pi}^{-1} 1\;dx + \int_{-1}^{1} (1-\lvert x \rvert) dx+ \int_{1}^{\pi} 1\;dx  \right] \\
    &= \frac{1}{2\pi}\int_{-\pi}^{\pi}1\;dx - \int_{-1}^{1} \lvert x \rvert\, dx \\
    &= 1- 2\int_{0}^{1} x \; dx \\
    &= 1- 1=0
\end{align*}

where we used the evenness of $\lvert x \rvert$ to rewrite $\int_{-1}^{1} \lvert x \rvert \,dx$ as $2\int_{0}^{1} x \; dx$. Thus, $a_0=0$. Now we perform a similar procedure to find $a_n$:

@@

Now inserting the coefficients $a_n$ we found above, we get the Fourier series:
$$f(x)=\frac{2}{\pi} \sum_{n=1}^{\infty}  \left(\frac{\sin{n}}{n}+ \frac{\cos{n}+1}{n^2} \right) \cos{ \left( nx \right)}$$

@@

The $\cos$ terms cancel and the $\sin$ terms add, yielding
$$\frac{x}{i} \sum_{n=-\infty}^{\infty}  \frac{(-1)^{n+1}} {i(1-n)}(- \sin{nx}) $$
Multiplying the two $i$s together cancels the negative sign, meaning the Fourier series becomes:
$$x \sin{x} =x \sum_{n=-\infty}^{\infty}  \frac{(-1)^{n+1}} {(1-n)}\sin{nx} $$

@@

>>>>

---
title: "MATH 245 Homework 4"
author: "Ruby Krasnow and Tommy Thach"
date: "2024-04-02"
output:
  pdf_document:
    extra_dependencies: ["tcolorbox", "xcolor"]
---

@@

Re-writing as $X''+\lambda X=0$, we will get the characteristic equation: $r^2+\beta^2=0$. Since our characteristic equation has complex roots $r= \pm i\beta$, our solutions take the form 
$$X(x)=A\sin{(\beta x)}+B\cos{(\beta x)}$$
Differentiating, we find that 
$$X'(x)=\beta A\cos{(\beta x)}-\beta B\sin{(\beta x)}$$
Now plugging in our initial condition $X'(0) = 0$, we get
$X'(0)=\beta A=0$. And since we are in a case where $\beta \neq 0$, this means $A=0$ so $X(x)=B\cos{(\beta x)}$. Now we use our boundary condition $X(l)=0$ to get $X(l)=B\cos{(\beta l)}=0$. If $B=0$, then $X(0)=0$ and this contradicts the definition of an eigenfunction. Therefore, we must have $\cos{(\beta l)}=0$, which can only occur if $\beta = \frac{(2n-1)\pi}{2l}$. Therefore, this case gives us eigenvalues

$$\lambda_n= \left(\frac{(2n-1)\pi}{2l}\right)^2 \quad n=1,2,3... $$
with eigenfunctions
$$X_n(x)=\cos{\left( \frac{(2n-1) \pi x}{2l}\right)}$$

@@

\underline{Case 2:} Zero eigenvalues, $\lambda=0$
$X''=0$ implies that $X(x)$ is of the form $Ax+B$. Now plugging in our initial condition $X(1) = 0$, we get $A+B=0$, which means $X(x)=Ax-A$. But from the boundary condition $X(e)=0$, we get $A(e-1)=0$ which is only possible if $A=0$ and accordingly $B=-A=0$. Therefore, there are no eigenfunctions $X(x)$ that satisfy the boundary conditions when $X''=0$ and hence no zero eigenvalues.

@@

$$X(1)=A+B=0 \quad \longrightarrow\quad A=-B$$
$$X(x)=Ax^{\beta }-Ax^{-\beta}$$
Then the other boundary condition gives $X(e)=Ae^{\beta}-Ae^{-\beta }=0 \quad\Rightarrow A(e^{\beta}-e^{-\beta })=0$. If we had $e^{\beta}-e^{-\beta }=0$, that would imply $\beta=-\beta\quad\Rightarrow\beta=0$, but we are in a case when we defined $\beta$ to be nonzero. So, we must have $A=0$, implying that again $X(x)=0$. Thus, this problem has only positive eigenvalues and their associated eigenfunctions as found in Case 1.

@@

$$X(1)=A\sin{(\beta)}+B\cos{(\beta)}=0$$
$$A\sin{(\beta)}=-B\cos{(\beta)}$$
Before dividing by $A$, we need to verify that $A \neq 0$. In this case, if $A = 0$,
then $X=B \cos{\beta x}$, so $X'(0)+X(0)=0+B=0$ implies $B=0$, a contradiction. Similarly, $\cos{\beta}\neq 0$, since if $\cos{\beta}=0$, then $\sin{\beta} \neq 0$ (sin and cos are not zero at the same time), and so
$A \sin{\beta}=-B\cos{\beta}=0 \quad\Rightarrow A=0$, leading to the same contradiction as before.

@@

$$ X'(0) + X(0) = \beta A-\beta B+A+B$$
$$0=A(\beta+1)+B(1-\beta)$$
$$-A(\beta+1)=B(1-\beta)$$
$$\frac{\beta+1}{1-\beta}=-\frac{B}{A}$$
Now turning to our other boundary condition,
$$X(1)=Ae^{\beta}+Be^{-\beta }=0$$
$$Ae^{\beta}=-Be^{-\beta}$$
$$-\frac{B}{A} = e^{2\beta}$$

We know $A \neq= 0$, since otherwise $X(1)=Be^{-\beta }=0$ implies $B = 0$, a contradiction.
Now you also can’t have $1-\beta = 0$, since $\beta = 1$ leads to $-A(1+1) = -2A = 0$. This justifies all the division so far.

This would give us the following equation:
$$\frac{\beta+1}{1-\beta}=e^{2\beta}$$
But the only time that this equality holds true is when $\beta=0$ (Fig. 2), but we are in the case when we defined $\beta$ to be nonzero, meaning that we have reached a contradiction and we cannot have any negative eigenvalues.

@@

Here, $l=\pi$. Let's start with $a_0$:
\begin{align*}
a_0 &=\frac{1}{2\pi} \int_{-\pi}^{\pi}f(x)dx \\
    &=\frac{1}{\pi} \int_{0}^{\pi}f(x)dx \\
    &= \frac{1}{\pi} \left(\int_{0}^{1}(1-\lvert x \rvert )\;dx +   \int_{1}^{\pi}\;dx \right) \\
    &= \frac{1}{\pi} \left(\int_{0}^{1}(1-x )\;dx +  (\pi-1) \right) \\
    &= \frac{1}{\pi} \left(\int_{0}^{1}(1-x )\;dx +  (\pi-1) \right) \\
    &= \frac{1}{\pi} \left((x-x^2/2) \Bigr\rvert_{0}^{1} +  (\pi-1) \right) \\
    &= \frac{1}{\pi} \left( \frac{1}{2} +  (\pi-1) \right) \\
    &= \frac{1}{\pi} \left( \pi-\frac{1}{2} \right) \\
\end{align*}

where we used the evenness of $\lvert x \rvert$ to rewrite $\int_{-\pi}^{\pi} f(x)dx \,dx$ as $2\int_{0}^{\pi} f(x)dx \; dx$. Thus, $a_0=\frac{1}{\pi} \left( \pi-\frac{1}{2} \right)$. Now we perform a similar procedure to find $a_n$:

@@

Now inserting the coefficients we found above, we get the Fourier series:
$$f(x)=\frac{1}{\pi} \left( \pi-\frac{1}{2} \right)+\frac{2}{\pi} \sum_{n=1}^{\infty}  \left(\frac{\sin{n}}{n}+ \frac{\cos{n}+1}{n^2} \right) \cos{ \left( nx \right)}$$

@@

The $\cos$ terms cancel and the $\sin$ terms add, yielding
$$\frac{x}{i} \sum_{n=-\infty}^{\infty}  \frac{(-1)^{n+1}} {i(1-n)}(- \sin{nx}) $$
Multiplying the two $i$s together cancels the negative sign, meaning the Fourier series becomes:
$$x \sin{x} =x \sum_{n=-\infty}^{\infty}  \frac{(-1)^{n+1}} {(1-n)}\sin{nx} $$
**Note that if $n = 1$, then $c_n = 0$, since you end up with the integral of $x$ over $(-\pi, \pi)$, which is zero. Hence the sum is properly over all terms with $n\neq 1$.

@@

If you wanted us to differentiate rather than integrate:
$$\frac{d}{dx} \sum^\infty_{k=1} (-1)^{k+1} \frac{\cos{kx}}{k}=\frac{d}{dx}\log{  \left(2 \cos{\left(\frac{x}{2}\right)}\right) }$$

\begin{align*}
\sum^\infty_{k=1} (-1)^{k}\sin{kx} &=\frac{-\sin{\frac{x}{2}}}{2 \cos{\frac{x}{2}}} \\
&= -\frac{1}{2}\tan{\frac{x}{2}}
\end{align*}

Thus, 
$$\sum^\infty_{k=1} (-1)^{k}\sin{kx}=-\frac{1}{2}\tan{\frac{x}{2}}$$

Note that this sum is actually divergent, since the terms don’t go to zero for any $x$ apart from a few special points (like $x = 0$).
